---
title: "crc_single_cell_revision_P1P5_honglin_20250715"
output: html_document
date: "2025-07-15"
---


process Honglin's single cell data and combine with our P1 P5 data and check for a few markers 




## check data 
```{r}

library(SingleCellExperiment)
library(DataSetDB)
library(gp.sa.core)
library(gp.sa.solo)
library(scater)
library(SingleR)
library(UCell)
library(dittoSeq)
library(parallel)
library(scran.chan)
library(BiocParallel)
library(celldex)
library(dsassembly)
ds = getDataset("DS000015552")
se = ds[[1]]
colD = colData(se)

akps = se[,which(colD$assignment %in% c("AKPSFULLAPCorganoid_DMSO","AKPSFULLAPCorganoid_DOX") & colD$demux_type == "singlet")]
colD_akps = colData(akps)
### these are all singlets 

table(colD_akps$Sample, colD_akps$assignment)
```
                  
                   AKPSFULLAPCorganoid_DMSO AKPSFULLAPCorganoid_DOX
  SAM24438240_rep1                     1879                     558
  SAM24438240_rep2                     1962                     573
  SAM24438240_rep3                     1971                     610
  

<!-- ```{r} -->
<!-- hash = altExp(akps,"hashing") -->
<!-- hash_assay = assay(hash) -->
<!-- ``` -->


<!-- ## assign cell type  -->
<!-- ```{r} -->
<!-- #  -->
<!-- # counts <- assay(akps, "counts") -->
<!-- # libsizes <- colSums(counts) -->
<!-- # size.factors <- libsizes/mean(libsizes) -->
<!-- # logcounts(akps) <- log2(t(t(counts)/size.factors) + 1) -->
<!-- # assayNames(akps) -->


<!-- library(scater) -->

<!-- akps = logNormCounts(akps) -->

<!-- library(scran) -->
<!-- top.gene = getTopHVGs(akps, n = 2000) -->


<!-- ``` -->





## process with gp.sa.solo
```{r}



## Change the directory location as needed:
proj.dir <- file.path("/gstore/project/crc_metastasis/revision/singleCell_colon/colon_sc/")



newProject(proj.dir, title="colon_scRNA",
           description="run basic gp.sa.solo pipeline, only singlets, 20250715")       
setwd(proj.dir)

rownames(akps) = rowData(akps)$Symbol
mito = grep("mt-", rownames(akps))
akps = addPerCellQCMetrics(akps, subsets = list(Mt = mito))
akps = akps[, which(akps$subsets_Mt_percent <25)]


output.akps <- runSingleCellBasic2(akps,
                                  symbol.field = "Symbol",
                                  qc.nmads = Inf, 
                                  rand.seed = 123 )

saveRDS(output.akps, "/gstore/project/crc_metastasis/revision/singleCell_colon/colon_sc/output.akps.rds")
```


```{r}
dittoDimPlot(output.akps$sce, "cluster",
             split.by = c("assignment","Sample"),
             do.label = T, labels.size = 2, labels.highlight = F)
```

```{r}
dittoDimPlot(output.akps$sce, "cluster",
             split.by = "assignment",
             do.label = T, labels.size = 2, labels.highlight = F)
```
```{r}

dittoDimPlot(output.akps$sce, "Epcam", 
             split.by = "assignment", size = 0.5, opacity = 0.5)+
    scale_color_viridis_c()
```


```{r}
dittoDimPlot(output.akps$sce, "cluster",
             split.by = "Sample",
             do.label = T, labels.size = 2, labels.highlight = F)
```


## annotate cluster 


automatic annotation with singleR

```{r}
library(SingleR)

imgen = ImmGenData()

pred.main <- SingleR(test = output.akps$sce, ref = imgen, assay.type.test='logcounts',labels = colData(imgen)$label.main, clusters = colData(output.akps$sce)$cluster, BPPARAM = MulticoreParam(workers = 4))

saveRDS(pred.main, "/gstore/project/crc_metastasis/revision/singleCell_colon/colon_sc/pred.main.Rds")

w.m = max.col(pred.main$scores)
max_label = colnames(pred.main$scores)[w.m]


pred_df = data.frame(cluster = as.factor(c(1:11)), SingleR = max_label, stringsAsFactors = F )


match_clu = data.frame(cluster = colData(output.akps$sce)$cluster, stringsAsFactors = F)%>%
  dplyr::left_join(pred_df, by = "cluster")

colData(output.akps$sce)$SingleR_max = match_clu$SingleR


saveRDS(output.akps, "/gstore/project/crc_metastasis/revision/singleCell_colon/colon_sc/output.sce.cluster.SingleR.Rds")


```


No surprise that these are all annotated to be epithelial cells because they are pure tumor organoid 

according to Honglin


## combine with Liming's organoid data 


```{r}
library(zellkonverter)

liming = readH5AD("/gstore/project/crc_metastasis/revision/data_agg_raw.h5ad")

library(HDF5Array)
out <- H5ADMatrix("/gstore/project/crc_metastasis/revision/data_agg_raw.h5ad")
m_out = as.matrix(out)


lim_sce =  SingleCellExperiment(list(counts=m_out),
    colData=DataFrame(colData(liming)),
    rowData=DataFrame(rowData(liming)))


saveRDS(lim_sce, "/gstore/project/crc_metastasis/revision/singleCell_colon/liming_sce.rds")

```


## readin colon sc organoid data and combine 

### these should all be epithellial cells 

```{r}
colon_sc = readRDS("/gstore/project/crc_metastasis/revision/singleCell_colon/colon_sc/output.sce.cluster.SingleR.Rds")
colon_sce = colon_sc$sce

lim_sce = readRDS("/gstore/project/crc_metastasis/revision/singleCell_colon/liming_sce.rds")
# 
# library(SingleCellExperiment)
# 
# output.merge <- gp.sa.solo2::runMerge(colon_sce, lim_sce)
# ### the abov
# 
# library(scRNAseq)
# sceG <- GrunPancreasData(ensembl=TRUE)
# sceM <- MuraroPancreasData(ensembl=TRUE)
# merged <- runMerge(sceM, sceG)


proj.dir <- file.path("/gstore/project/crc_metastasis/revision/singleCell_colon/colon_sc_liming_merge/")



newProject(proj.dir, title="colon_liming_scRNA",
           description="run merge gp.sa.solo pipeline,liming's organoid and Honglin's organoid, 20250716")       
setwd(proj.dir)

output.merge <- runSingleCellMerge2(colon_sce,
                                    lim_sce,
                                    batch = "study",
                                  rand.seed = 123 )
### the above has a bug will work on the markdown file direclty 





library(scRNAseq)
sceZ <- ZeiselBrainData(ensembl=TRUE)
sceT <- TasicBrainData(ensembl=TRUE)


```







