---
title: "crc_revision_test_Qin_signatures_20250527"
output: html_document
date: "2025-05-27"
---


## read in Qin et al signatures 
https://www.cell.com/cell/pdf/S0092-8674(23)01221-7.pdf
```{r}

library(dplyr)
library(data.table)
library(magrittr)
library(readxl)

qsig = read_excel("/gstore/project/crc_metastasis/revision/Qin_et_al/mmc3.xlsx", sheet = 2,skip = 4)

qsig_sig = as.data.frame(qsig[,c(1,2)])%>%
  na.omit()%>%
  dplyr::mutate(gene_set = paste0("signature_", ANNOTATION))
colnames(qsig_sig)[1] = "gene"

qsig_de = as.data.frame(qsig[,c(4,5)])%>%
  na.omit()%>%
  dplyr::mutate(gene_set = paste0("DEclust_", CLUSTER))
colnames(qsig_de)[1] = "gene"

qpaths = rbind(qsig_sig[,c(1,3)], qsig_de[,c(1,3)])%>%
  dplyr::select(gene_set, gene)%>%
  dplyr::filter(gene_set %in% c("DEclust_proCSC","DEclust_CSC","DEclust_revCSC",
                                "signature_proCSC","signature_CSC","signature_revCSC"))


### add the RSC and CBC signatures 
rsc = fread("/gstore/home/lix335/resources/rsc_cbc.tsv", 
            stringsAsFactors = F, data.table = F)
colnames(rsc) = c("gene_set","gene")

lit_sets = rbind(qpaths, rsc)

```



## perform GSEA 


```{r}
library(clusterProfiler)


res_df = fread("/gstore/project/crc_metastasis/P5_P1_revisit_20240801/res_P5_P1.tsv", stringsAsFactors = F, data.table = F)%>%
  dplyr::arrange(symbol)


geneList = res_df$t
names(geneList) = res_df$symbol
geneList = sort(geneList, decreasing = T)
  

set.seed(47382686)

#y <- GSEA(geneList, TERM2GENE = qpaths, pvalueCutoff = 1, minGSSize = 4, maxGSSize = 700, seed = T)
y <- GSEA(geneList, TERM2GENE = lit_sets, pvalueCutoff = 1, minGSSize = 4, maxGSSize = 700, seed = T)


# 
# set.seed(123)
# 
# y1 <- GSEA(geneList, TERM2GENE = qpaths, pvalueCutoff = 1, minGSSize = 4, maxGSSize = 700, seed = T)

```

## create a plot 


```{r}
library(enrichplot)
dotplot(y)
```

```{r}
library(ggplot2)

gsea_res = y@result%>%
  dplyr::mutate(regulation = ifelse(NES>0, "up","down"))

b= ggplot(gsea_res, aes(x = reorder(ID, NES),  y = NES, 
                       color = regulation, 
                       fill = sign(NES)*(-log10(p.adjust))))+
  geom_bar(stat = "identity")+
  scale_fill_gradient2(low = "blue", mid = "white", high = "red")+
  scale_color_manual(values = c("blue","red"))+
  coord_flip()+
  theme(axis.text.y = element_text(size = 5))+
  theme_bw()+
  labs(x = NULL)


b
```



##heatmaps for individual genes in P1 and P5


```{r}


expr = fread("/gstore/project/crc_metastasis/P5_P1_revisit_20240110/voom-results/filter_lcpm_expression_correct.tsv", 
            stringsAsFactors = F)


res_df = fread("/gstore/project/crc_metastasis/P5_P1_revisit_20240801/res_P5_P1.tsv", stringsAsFactors = F, data.table = F)%>%
  dplyr::arrange(symbol)
```


## individual signature detailed heatmaps
```{r}


library(matrixStats)
library(circlize)
library(ComplexHeatmap)

sets = unique(qpaths$gene_set)
for(i in 1:length(sets))
{
  
  genes = qpaths$gene[which(qpaths$gene_set == sets[i])]
  

qin_expr = expr%>%
  dplyr::filter(symbol %in% genes)

expr_mat = as.matrix(qin_expr[, c(3:9)])
colnames(expr_mat) = colnames(qin_expr)[3:9]
rownames(expr_mat) = qin_expr$symbol

### standardize rowwise 


r_m = rowMeans(expr_mat)
r_sd = rowSds(expr_mat)
stan_expr = sweep(expr_mat, 1, r_m)
stan_expr = sweep(stan_expr, 1, r_sd, FUN = "/")

col_fun_expr = colorRamp2(c(-2, 0, 2), c("blue","white", "red"))


ta = HeatmapAnnotation(group =c(rep("P1",4), rep("P5",3)),
                        col = list(group = c("P1" = "#2D8686",
                                            "P5" = "#862D2D")))


square_cal = function(n)
{
  if(n > 10)
  {
      c = ceiling(log10(n))
  p = n/(10**c)
  }else{
      c = ceiling(log10(n))
  p = n/(100**c)
  
  }

  if(n == 1)
  {
    p = 0.01
  }
  
  return(p)
  
}
hei = unit(square_cal(nrow(as.matrix(stan_expr)))*50, "cm")


h1 = Heatmap(as.matrix(stan_expr),
             name = "norm.exp.",
             column_title = sets[i],
        col = col_fun_expr,
        height = hei, 
        width = hei*(ncol(stan_expr)/nrow(stan_expr)),
        row_names_gp = gpar(fontsize = 6),
        column_names_gp = gpar(fontsize = 4),
        rect_gp = gpar(col = "black", lwd = 0.5),
        top_annotation = ta, 
        cluster_rows = T, cluster_columns = F,
        column_split = factor(c(rep("P1",4), rep("P5",3)), levels = c("P1", "P5")))


lf_df = rbindlist(lapply(1:nrow(qin_expr), function(x) {
  this_gene = qin_expr$symbol[x]

    gp15 = res_df%>%
          dplyr::filter(symbol == this_gene)

    
  df = data.frame( symbol = this_gene,
                   lf_p15 = NA, 
                   fdr_p15 = NA, 
                   sig_p15 = "",
                  stringsAsFactors = F)
   if(nrow(gp15) >0)
  {
        df$lf_p15 = gp15$LogFC[1]
        df$fdr_p15 =  gp15$FDR[1]
             if(gp15$FDR[1] <= 0.05)
          df$sig_p15 = "*"

  }
  
  return(df)
  
}))

sig_mat = as.matrix(lf_df[,4])

lf_mat = as.matrix(lf_df[,2])
colnames(lf_mat) =  "P5vsP1"

rownames(lf_mat) = lf_df$symbol

col_fun = colorRamp2(c(-2, 0,2), c("blue", "white", "red"))




h2 = Heatmap(as.matrix(lf_mat),
             name = "Log2 FC",
        col = col_fun,
        height = hei, 
        width = hei*(ncol(lf_mat)/nrow(lf_mat)),
         cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%s", sig_mat[i, j]), x, y, gp = gpar(fontsize = 8))
},
        rect_gp = gpar(col = "black", lwd = 0.5),
        column_names_gp = gpar(fontsize = 4),
        row_names_gp = gpar(fontsize = 6),
        cluster_rows = F, cluster_columns = F)


this_height = 7+ (nrow(lf_mat)-10)*0.1

pdf(paste0("/gstore/project/crc_metastasis/P5_P1_revisit_20240801/revision_Qin_", sets[i],"_heatmap_20250617.pdf"),
    useDingbats = F, height =this_height)

draw(h1+h2)

dev.off()

}



### just the two rsc cbc sets 


sets = c("rsc","cbc")
for(i in 1:length(sets))
{
  genes = lit_sets$gene[which(lit_sets$gene_set == sets[i])]
  

qin_expr = expr%>%
  dplyr::filter(symbol %in% genes)

expr_mat = as.matrix(qin_expr[, c(3:9)])
colnames(expr_mat) = colnames(qin_expr)[3:9]
rownames(expr_mat) = qin_expr$symbol

### standardize rowwise 


r_m = rowMeans(expr_mat)
r_sd = rowSds(expr_mat)
stan_expr = sweep(expr_mat, 1, r_m)
stan_expr = sweep(stan_expr, 1, r_sd, FUN = "/")

col_fun_expr = colorRamp2(c(-2, 0, 2), c("blue","white", "red"))


ta = HeatmapAnnotation(group =c(rep("P1",4), rep("P5",3)),
                        col = list(group = c("P1" = "#2D8686",
                                            "P5" = "#862D2D")))



h1 = Heatmap(as.matrix(stan_expr),
             name = "norm.exp.",
             column_title = sets[i],
        col = col_fun_expr,
        #height = hei, 
        #width = hei*(ncol(stan_expr)/nrow(stan_expr)),
        row_names_gp = gpar(fontsize = 6),
        column_names_gp = gpar(fontsize = 4),
        rect_gp = gpar(col = "black", lwd = 0.5),
        top_annotation = ta, 
        cluster_rows = T, cluster_columns = F,
        column_split = factor(c(rep("P1",4), rep("P5",3)), levels = c("P1", "P5")))


lf_df = rbindlist(lapply(1:nrow(qin_expr), function(x) {
  this_gene = qin_expr$symbol[x]

    gp15 = res_df%>%
          dplyr::filter(symbol == this_gene)

    
  df = data.frame( symbol = this_gene,
                   lf_p15 = NA, 
                   fdr_p15 = NA, 
                   sig_p15 = "",
                  stringsAsFactors = F)
   if(nrow(gp15) >0)
  {
        df$lf_p15 = gp15$LogFC[1]
        df$fdr_p15 =  gp15$FDR[1]
             if(gp15$FDR[1] <= 0.05)
          df$sig_p15 = "*"

  }
  
  return(df)
  
}))

sig_mat = as.matrix(lf_df[,4])

lf_mat = as.matrix(lf_df[,2])
colnames(lf_mat) =  "P5vsP1"

rownames(lf_mat) = lf_df$symbol

col_fun = colorRamp2(c(-2, 0,2), c("blue", "white", "red"))




h2 = Heatmap(as.matrix(lf_mat),
             name = "Log2 FC",
        col = col_fun,
        #height = hei, 
        #width = hei*(ncol(lf_mat)/nrow(lf_mat)),
         cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%s", sig_mat[i, j]), x, y, gp = gpar(fontsize = 8))
},
        rect_gp = gpar(col = "black", lwd = 0.5),
        column_names_gp = gpar(fontsize = 4),
        row_names_gp = gpar(fontsize = 6),
        cluster_rows = F, cluster_columns = F)


this_height = 7+ (nrow(lf_mat)-10)*0.05

pdf(paste0("/gstore/project/crc_metastasis/P5_P1_revisit_20240801/revision_Qin_", sets[i],"_heatmap_20250619.pdf"),
    useDingbats = F, height =this_height)

draw(h1+h2)

dev.off()

}



```


## correlation between these signature scores and the mapk signature scores 
### a note: it is necessary to do normalization in the eigenWeightedMean calculation in order to compare with the GSEA results 
###https://rdrr.io/github/lianos/multiGSEA/man/eigenWeightedMean.html
```{r}

library(multiGSEA)


mlist = c("Spred1","Spred2", "Ets","Spry2", "Spry4", "Etv1", "Etv4", "Etv6", "Dusp4","Dusp6")
wlist = c("Lgr5", "Axin2", "Ascl2", "Slc12a2", "Gkn33p", "Nkd1", "Wif1", "Smoc2")

f_df = fread("/gstore/home/lix335/crc_metastasis_ATAC/motif_peak_BETA_network/top6_spacer_leftcf_motifs_signature_highEdge_sets8_merge_sig0.0001_20241103.tsv", 
             stringsAsFactors = F)

colnames(f_df)[1] = "gene_set"

c_sets = data.frame(gene_set = c(rep("c_mapk", length(mlist)), rep("c_wnt", length(wlist))), 
                    gene = c(mlist, wlist), stringsAsFactors = F)

all_sets = rbind(f_df,  c_sets,lit_sets)

## calculate a score for each of these signatures 
lcpm = as.matrix(expr[, -c(1:2)])
rownames(lcpm) = expr$symbol


all_sets_names = unique(all_sets$gene_set)

score_df= rbindlist(lapply(1:length(all_sets_names), function(i) {
  
  this_set = all_sets_names[i]
  this_genes = all_sets$gene[which(all_sets$gene_set == this_set)]
  
  this_genes = intersect(rownames(lcpm), this_genes)
  
  set.seed(123)

  this_score<- eigenWeightedMean(lcpm[this_genes,], center = T, scale = T, normalize = T,all.x = lcpm)$score

  df = data.frame(signature = this_set, score = this_score, ID = names(this_score),stringsAsFactors = F)
  
  return(df)
}))



# 
score_df2 = rbindlist(lapply(1:length(all_sets_names), function(i) {
  this_set = all_sets_names[i]
  this_genes = all_sets$gene[which(all_sets$gene_set == this_set)]

  this_genes = intersect(rownames(lcpm), this_genes)
  this_score<- eigenWeightedMean(lcpm[this_genes,], center = T, scale = T)$score

  df = data.frame(signature = this_set, score = this_score, ID = names(this_score),stringsAsFactors = F)

  return(df)
}))

# #
# score_df1 = rbindlist(lapply(1:length(all_sets_names), function(i) {
#   this_set = all_sets_names[i]
#   this_genes = all_sets$gene[which(all_sets$gene_set == this_set)]
#
#   this_genes = intersect(rownames(lcpm), this_genes)
#   this_score<- apply(lcpm[this_genes,],2,mean)
#
#   df = data.frame(signature = this_set, score = this_score, ID = names(this_score),stringsAsFactors = F)
#
#   return(df)
# }))


score_mat2 = score_df2%>%
  tidyr::pivot_wider(names_from = ID, values_from = score)
s_mat2 = as.matrix(score_mat2[, -1])
rownames(s_mat2) = score_mat2$signature


```

### plot a heatmap 

```{r}
library(tidyr)

score_mat = score_df%>%
  tidyr::pivot_wider(names_from = ID, values_from = score)
s_mat = as.matrix(score_mat[, -1])
rownames(s_mat) = score_mat$signature

## also perform a test on the difference between the scores 
score_test = rbindlist(lapply(1:nrow(s_mat), function(x) {
 p1_scores = as.numeric(s_mat[x, 1:4])
 p5_scores = as.numeric(s_mat[x, 5:7])
 
 score_diff = median(p5_scores) - median(p1_scores) 
 
 s_test = wilcox.test(p5_scores, p1_scores)
 score_p = s_test$p.value
 
 df = data.frame(score_diff, score_p)
 
}))



### standardize rowwise 


r_m = rowMeans(s_mat)
r_sd = rowSds(s_mat)
stan_expr = sweep(s_mat, 1, r_m)
stan_expr = sweep(stan_expr, 1, r_sd, FUN = "/")

col_fun_expr = colorRamp2(c(-1.5, 0, 1.5), c("blue","white", "red"))


ta = HeatmapAnnotation(group =c(rep("P1",4), rep("P5",3)),
                        col = list(group = c("P1" = "#2D8686",
                                            "P5" = "#862D2D")))


square_cal = function(n)
{
  if(n > 10)
  {
      c = ceiling(log10(n))
  p = n/(10**c)
  }else{
      c = ceiling(log10(n))
  p = n/(100**c)
  
  }

  if(n == 1)
  {
    p = 0.01
  }
  
  return(p)
  
}
hei = unit(square_cal(nrow(as.matrix(stan_expr)))*50, "cm")


h1 = Heatmap(as.matrix(stan_expr),
             name = "normalized score",
             column_title = "signature scores",
        col = col_fun_expr,
        height = hei, 
        width = hei*(ncol(stan_expr)/nrow(stan_expr)),
        row_names_gp = gpar(fontsize = 6),
        column_names_gp = gpar(fontsize = 4),
        rect_gp = gpar(col = "black", lwd = 0.5),
        top_annotation = ta, 
        cluster_rows = T, cluster_columns = F,
        column_split = factor(c(rep("P1",4), rep("P5",3)), levels = c("P1", "P5")))






sig_mat = as.matrix(round(score_test[,2],3))

lf_mat = as.matrix(score_test[,1])
colnames(lf_mat) =  "P5vsP1"

rownames(lf_mat) = rownames(s_mat)

col_fun = colorRamp2(c(-3, 0,3), c("blue", "white", "red"))




h2 = Heatmap(as.matrix(lf_mat),
             name = "Score median diff",
        col = col_fun,
        height = hei, 
        width = hei*(ncol(lf_mat)/nrow(lf_mat)),
         cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%s", sig_mat[i, j]), x, y, gp = gpar(fontsize = 5))
},
        rect_gp = gpar(col = "black", lwd = 0.5),
        column_names_gp = gpar(fontsize = 4),
        row_names_gp = gpar(fontsize = 6),
        cluster_rows = F, cluster_columns = F)



pdf(paste0("/gstore/project/crc_metastasis/P5_P1_revisit_20240801/revision_Qin_rsc_cbc_US_sigantures_heatmap_pvalue_20250619.pdf"),
    useDingbats = F, height =7)

draw(h1+h2)

dev.off()


```

### create correlation between these signature scores 


```{r}
library(reshape2)


reorder_cormat <- function(cormat){
# Use correlation between variables as distance
dd <- as.dist((1-cormat)/2)
hc <- hclust(dd)
cormat <-cormat[hc$order, hc$order]
}


sig_cor = reorder_cormat(cor(t(s_mat)))
# 
# melted_cormat <- reshape2::melt(sig_cor)
# 
# ggplot(data = melted_cormat, aes(x=Var1, y=Var2, fill=value)) + 
#   geom_tile()+
#     scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, limits = c(-1, 1))+
#     theme(axis.text.x = element_text(angle = 90, hjust = 1))
# 





col_fun_cor = colorRamp2(c(-1, 0, 1), c("blue","white", "red"))



square_cal = function(n)
{
  if(n > 10)
  {
      c = ceiling(log10(n))
  p = n/(10**c)
  }else{
      c = ceiling(log10(n))
  p = n/(100**c)
  
  }

  if(n == 1)
  {
    p = 0.01
  }
  
  return(p)
  
}
hei = unit(square_cal(nrow(as.matrix(sig_cor)))*50, "cm")


h1 = Heatmap(as.matrix(sig_cor),
             name = "correlation",
             #column_title = "signature scores",
        col = col_fun_cor,
        height = hei, 
        width = hei,
        row_names_gp = gpar(fontsize = 6),
        row_names_side = "left",
        column_names_gp = gpar(fontsize = 6),
        rect_gp = gpar(col = "black", lwd = 0.5),
        #top_annotation = ta, 
        cluster_rows = F, cluster_columns = F)


pdf(paste0("/gstore/project/crc_metastasis/P5_P1_revisit_20240801/revision_Qin_rsc_cbc_US_sigantures_heatmap_correlation_20250619.pdf"),
    useDingbats = F, height =7)

draw(h1)

dev.off()

```





I need to understand why after normalization, the direction changed 

```{r}



r_m2 = rowMeans(s_mat2)
stan_expr2 = sweep(s_mat2, 1, r_m2)

col_fun_expr = colorRamp2(c(-1.5, 0, 1.5), c("blue","white", "red"))



h2 = Heatmap(as.matrix(stan_expr2),
             name = "normalized score",
             column_title = "signature scores",
        col = col_fun_expr,
        height = hei, 
        width = hei*(ncol(stan_expr)/nrow(stan_expr)),
        row_names_gp = gpar(fontsize = 6),
        column_names_gp = gpar(fontsize = 4),
        rect_gp = gpar(col = "black", lwd = 0.5),
        top_annotation = ta, 
        cluster_rows = T, cluster_columns = F,
        column_split = factor(c(rep("P1",4), rep("P5",3)), levels = c("P1", "P5")))

boxplot(lcpm)

```



Conclusion:

using Normalize = T, I subtract a mean level from P1 and P5 samples, and because the mean levels in P1 are larger, so after subtraction, the scores of P1 gets smaller. 









