---
title: "crc_revision_allPassages_WES_20250529"
output: html_document
date: "2025-05-29"
---
## reprocessed WES somatic calling data for all stages 
SRA25044

```{r}

library(data.table)
library(dplyr)
library(magrittr)
library(ggplot2)
library(VariantAnnotation)
library(Vennerable)



```

## Study the quality of the sequencing between passages 
P1P2P5 and P3P4 are from different batches


```{r}

samIDs = c("SAM24400842","SAM24400845","SAM24488704","SAM24488703","SAM24400853")
ps = c("P1","P2","P3","P4","P5")



qc_align_files = paste0("/gstore/data/genomics/congee_rest_runs/68377ab0f719730fbb8e88db/",
                  samIDs, "/croo_output/qc/", samIDs, ".GRCm38.exome-T.alignment_summary_metrics")
get_stats = rbindlist(lapply(1:length(samIDs), function(x) {
  f = fread(qc_align_files[x], stringsAsFactors = F, data.table = F)%>%
    dplyr::filter(CATEGORY == "PAIR")%>%
    dplyr::select(-LIBRARY)%>%
    dplyr::mutate(passage = ps[x])%>%
    unique()%>%
    dplyr::select(SAMPLE, passage, everything())
  
  
  return(f)
  
}))

```



## plot the stats 

```{r}

tr = ggplot(get_stats, aes(x = passage, y = TOTAL_READS))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))


pct_pf = ggplot(get_stats, aes(x = passage, y = PCT_PF_READS))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))




pf_reads_aligned = ggplot(get_stats, aes(x = passage, y = PF_READS_ALIGNED ))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))


pct_pf_reads_aligned = ggplot(get_stats, aes(x = passage, y = PCT_PF_READS_ALIGNED ))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))


reads_aligned_pairs = ggplot(get_stats, aes(x = passage, y = READS_ALIGNED_IN_PAIRS ))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))


pct_reads_aligned_pairs = ggplot(get_stats, aes(x = passage, y = PCT_READS_ALIGNED_IN_PAIRS ))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))


pf_mismatch_rate = ggplot(get_stats, aes(x = passage, y = PF_MISMATCH_RATE ))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))



pf_indel_rate = ggplot(get_stats, aes(x = passage, y = PF_INDEL_RATE ))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))



library(patchwork)

tr + pct_pf +pf_reads_aligned+ pct_pf_reads_aligned+ reads_aligned_pairs+ 
  pct_reads_aligned_pairs+ pf_mismatch_rate+ pf_indel_rate + plot_layout(ncol = 2)

```

## quality distribution qc 

```{r}

qc_dist_files = paste0("/gstore/data/genomics/congee_rest_runs/68377ab0f719730fbb8e88db/",
                  samIDs, "/croo_output/qc/", samIDs, ".GRCm38.exome-T.quality_distribution_metrics")
get_stats = rbindlist(lapply(1:length(samIDs), function(x) {
  f = fread(qc_dist_files[x], stringsAsFactors = F, data.table = F, header = T)
  
  if(grepl("#", colnames(f)[1]))
  {
    fnew = f[-1,]
    colnames(fnew) = f[1,]
  }else{
    fnew = f
  }
  f_data = fnew%>%
        dplyr::mutate(SAMPLE =samIDs[x], passage = ps[x])
  
  
  return(f_data)
  
}))

get_stats = get_stats%>%
  dplyr::mutate(QUALITY = factor(QUALITY, levels = c("2","3","4","5","6","10","20","30")))%>%
  dplyr::mutate(COUNT_OF_Q = as.numeric(COUNT_OF_Q))

### plot out 


ggplot(get_stats, aes(QUALITY, log2(COUNT_OF_Q)))+
  geom_bar(stat = "identity")+
  facet_wrap(~passage)

```

```{r}

ggplot(get_stats, aes(QUALITY, COUNT_OF_Q))+
  geom_bar(stat = "identity")+
  facet_wrap(~passage, nrow = 1)+
  theme_bw()

```
## insert size 

```{r}

qc_insert_files = paste0("/gstore/data/genomics/congee_rest_runs/68377ab0f719730fbb8e88db/",
                  samIDs, "/croo_output/qc/", samIDs, ".GRCm38.exome-T.insert_size_metrics")
get_stats = rbindlist(lapply(1:length(samIDs), function(x) {
  f = fread(qc_insert_files[x], stringsAsFactors = F, data.table = F, header = T, skip = 11)
  
  if(grepl("#", colnames(f)[1]))
  {
    fnew = f[-1,]
    colnames(fnew) = f[1,]
  }else{
    fnew = f
  }
  f_data = fnew[,c(1,2)]%>%
        dplyr::mutate(SAMPLE =samIDs[x], passage = ps[x])
  
  colnames(f_data)[2] = "count"
  
  return(f_data)
  
}))


ggplot(get_stats, aes(insert_size, count))+
  geom_bar(stat = "identity")+
  facet_wrap(~passage, nrow = 1)+
  theme_bw()

```

```{r}


qc_dup_files = paste0("/gstore/data/genomics/congee_rest_runs/68377ab0f719730fbb8e88db/",
                  samIDs, "/croo_output/qc/", samIDs, ".GRCm38.exome-T.duplicate_metrics")

get_stats = rbindlist(lapply(1:length(samIDs), function(x) {
  f = fread(qc_dup_files[x], stringsAsFactors = F, data.table = F)%>%
    dplyr::select(-LIBRARY)%>%
    dplyr::mutate(passage = ps[x])%>%
    unique()
  
  return(f)
  
}))



unpaired_reads = ggplot(get_stats, aes(x = passage, y = UNPAIRED_READS_EXAMINED))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))


paired_reads = ggplot(get_stats, aes(x = passage, y = READ_PAIRS_EXAMINED))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))


unpaired_reads_dup = ggplot(get_stats, aes(x = passage, y = UNPAIRED_READ_DUPLICATES))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))

paired_reads_dup = ggplot(get_stats, aes(x = passage, y = READ_PAIR_DUPLICATES))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))

pct_dup = ggplot(get_stats, aes(x = passage, y = PERCENT_DUPLICATION))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))

lib_size = ggplot(get_stats, aes(x = passage, y = ESTIMATED_LIBRARY_SIZE))+
  geom_bar(stat = "identity")+
  theme_bw()+
  theme(axis.title.y = element_text(size = 7))






library(patchwork)

paired_reads + unpaired_reads + paired_reads_dup + unpaired_reads_dup + pct_dup + lib_size +plot_layout(ncol = 2)


```





## gather the list of mutations of interest 


```{r}

indir.lst  = paste0("/gstore/data/genomics/congee_rest_runs/68377ab0f719730fbb8e88db/", samIDs,"/croo_output")
cwd <- "/gstore/project/crc_metastasis/revision/P1P2P3P4P5_WES"
grp <- "P0_2"


results <- lapply(indir.lst, function(indir){
  #indir = indir.lst[1]
  message(indir)
  SAMID <- unlist(strsplit(indir, "/"))[7]

  f1 <- list.files(indir, pattern="snvs.*out", full.names = TRUE)
  f2 <- list.files(indir, pattern="mutect.*out", full.names = TRUE)

  fl <- c(f1,f2)
  fl = grep(".tbi", fl, value = T, invert = T)

  vcf.l <- lapply(fl, function(f){
    readVcf(f)
  })

  names(vcf.l) <- sapply(strsplit(basename(fl), "_"), "[[", 2)


vcf.processed.l <- lapply(names(vcf.l), function(x){

    if(x == "lofreq"){
      message("processing lofreq")
      vcf.processed <- process.lofreq(vcf.l[[x]])
    } else if(x == "strelka"){
      message("processing strelka")
      vcf.processed <- process.strelka(vcf.l[[x]])
    } else if( x == "mutect2"){
      message("processing mutect")
      vcf.processed <- process.mutect2(vcf.l[[x]], pass.only = FALSE)
    }

    return(vcf.processed)
  })

  names(vcf.processed.l) <- names(vcf.l)

  ################
  # plot venn of 
  vcf.l.names <- lapply(vcf.processed.l, function(x){
    x$ID
  })

  v <- Venn(vcf.l.names)
  pdf(file.path(cwd, sprintf("figures/%s.somaticcalleroverlap.%s.pdf", SAMID, grp)))
  plot(v)
  dev.off()

  vcf.basic.l <- lapply(vcf.processed.l, function(x){
    x[,1:9]
  })

  vcf.basic.bind <- do.call(rbind, vcf.basic.l)
  ID.tab <- table(vcf.basic.bind$ID)
  vcf.basic.bind.uniq <- unique(vcf.basic.bind)
  vcf.basic.bind.uniq$n.methods <- ID.tab[match(vcf.basic.bind.uniq$ID, names(ID.tab))]

  vcf.basic.bind.uniq$AF.lofreq <- vcf.processed.l$lofreq[match(vcf.basic.bind.uniq$ID, vcf.processed.l$lofreq$ID),"AF"]
  vcf.basic.bind.uniq$AF.strelka <- vcf.processed.l$strelka[match(vcf.basic.bind.uniq$ID, vcf.processed.l$strelka$ID),"AF"]
  vcf.basic.bind.uniq$AF.mutect <- vcf.processed.l$mutect2[match(vcf.basic.bind.uniq$ID, vcf.processed.l$mutect2$ID),"AF"]
  vcf.basic.bind.uniq$AF.mean <- rowMeans(vcf.basic.bind.uniq[,c("AF.lofreq","AF.strelka", "AF.mutect")], na.rm = TRUE)

  vcf.basic.bind.uniq$CSQ.lofreq <- vcf.processed.l$lofreq[match(vcf.basic.bind.uniq$ID, vcf.processed.l$lofreq$ID),"CSQ"]
  vcf.basic.bind.uniq$CSQ.strelka <- vcf.processed.l$strelka[match(vcf.basic.bind.uniq$ID, vcf.processed.l$strelka$ID),"CSQ"]
  vcf.basic.bind.uniq$CSQ.mutect <- vcf.processed.l$mutect2[match(vcf.basic.bind.uniq$ID, vcf.processed.l$mutect2$ID),"CSQ"]

  vcf.basic.bind.uniq <- vcf.basic.bind.uniq %>% arrange(desc(n.methods), desc(AF.mean))

  return(vcf.basic.bind.uniq)

})



names(results) <- sapply(strsplit(indir.lst, "/"), "[[",7)

saveRDS(results, file.path(cwd, sprintf("results/NGS6581.%s.somatic_calls.rds",grp)))

effects <- apply(results$SAM24488703[, c("CSQ.lofreq", "CSQ.strelka", "CSQ.mutect")], 1 , paste, collapse=" | ")
effects.l <- strsplit(effects, " \\| ")
sapply(seq_along(effects.l), function(i){
  message(i)
  x <- effects.l[[i]][grepl(",",effects.l[[i]])]
  sapply(strsplit(x, ","), "[[", 2)
}) %>% unlist() %>% table() %>% sort()

effects.interest <- c("missense_variant", "NMD_transcript_variant", "frameshift_variant", "start_lost", "stop_gained")

################
# select results that only have the specific effects
results.sel <- lapply(names(results), function(samid){

  r <- results[[samid]]

  r$SAMID <- samid

  effects <- apply(r[, c("CSQ.lofreq", "CSQ.strelka", "CSQ.mutect")], 1 , paste, collapse=" | ")
  r$effects.merged  <- effects

  r$use.effect <- sapply(seq_along(effects), function(i){
    any(sapply(effects.interest, function(x){
      grepl(x, effects[[i]])
    })
    )
  })

  r.sel <- r %>% filter(AF.mean > 0.39 & use.effect & n.methods > 1)
  return(r.sel)
})


results.sel.bind <- do.call(rbind, results.sel)

library(tidyverse)
r <- results.sel.bind %>% dplyr::select(ID, effects.merged, SAMID, AF.mean) %>% unique()%>%spread(SAMID, AF.mean) %>%
  mutate(AF.clonemean = rowMeans(dplyr::select(.,starts_with("SAM")), na.rm = TRUE),
         n.AF.present = rowSums(!is.na(dplyr::select(.,starts_with("SAM"))))) %>%
  arrange(desc(n.AF.present), desc(AF.clonemean))
  View()

saveRDS(r, file.path(cwd, sprintf("results/NGS6581.%s.somatic_calls_summary.rds", grp)))

```

Somatic_analyze_vafs.R
##find the interested mutations from the original bam files 


```{r}
#library(VariantFreq)
library(igis)
library(pheatmap)

igisRConfFile <<- file.path(tempdir(), "igisR.conf")

############
# load positions
r <- readRDS(file.path(cwd, sprintf("results/NGS6581.%s.somatic_calls_summary.rds", grp)))

rpos <- do.call(rbind,strsplit(r$ID, "[_:]+"))
rpos = unique(rpos)
# 
# effects.sep <- lapply(strsplit(r$effects.merged, " \\| "), function(x){
#   x <- x[x!="NA"]
#   paste(unique(sapply(strsplit(x, ","), "[[", 1)), collapse=",")
#   })

rnucs <- data.frame(do.call(rbind,strsplit(rpos[,3],"/")))
colnames(rnucs) <- c("ref", "alt")
rnucs$ID <- paste(rpos[,1], rpos[,2], sep=":")
#rnucs$gene <- effects.sep


posExample <- GenomicRanges::GRanges(rpos[,1],
                                     IRanges::IRanges(start = as.numeric(rpos[,2]), width = 1))
mcols(posExample) <- rnucs
#posExample <- unique(posExample)

#############
# get bamfiles
files.bam <- paste0("/gstore/data/genomics/congee_rest_runs/68377ab0f719730fbb8e88db/",
                  samIDs, "/croo_output/", samIDs, ".GRCm38.exome-T.bam")
names(files.bam) <- sapply(strsplit(basename(files.bam), "\\."), "[[", 1)


#############
# get nucleotide frequencies
freqs <- getAllelesFromBamSimple(posExample, files.bam)


freqs.sel <- lapply(freqs, function(f){
  #f = freqs[[1]]
  mc <- mcols(posExample)
  f <- cbind(mc, f[match(mc$ID,f$position),])
  f$fracAlt <- sapply(1:nrow(f),function(i){
    f[i, paste0("frac",f[i,"alt"])]
  })
  f$ID_all = paste0(f$ID, "_",f$ref,"/",f$alt)
  return(setNames(f$fracAlt, f$ID_all))
})
names(freqs.sel) <- names(freqs)

freqs.sel.bind <- do.call(cbind, freqs.sel)



freqs.sel.reads <- lapply(freqs, function(f){
  #f = freqs[[1]]
  mc <- mcols(posExample)
  f <- cbind(mc, f[match(mc$ID,f$position),])
  f$fracAlt <- sapply(1:nrow(f),function(i){
    f[i, paste0("frac",f[i,"alt"])]
  })
  f$ID_all = paste0(f$ID, "_",f$ref,"/",f$alt)
  return(setNames(f$numberOfReads, f$ID_all))
})
names(freqs.sel.reads) <- names(freqs)

freqs.sel.reads.bind <- do.call(cbind, freqs.sel.reads)

###############
# # combine frequencies
# NGS.proj.info <- gneDB::ngsSampleMetadata("NGS6581")
# NGS.proj.info$name <- gsub("\\(","",gsub("\\)","",gsub("-","_",gsub(" ", "_",NGS.proj.info$INVENTORY_SAMPLE_NAME))))
# 
# d4plot <- freqs.sel.bind
# colnames(d4plot) <- NGS.proj.info$name[match(colnames(d4plot), NGS.proj.info$SAMPLE_ID_DISPLAY_VALUE)]
# rownames(d4plot) <- paste(posExample$ID, posExample$gene, sep="_")
# colnames(d4plot[,c(14,2,3,4,5,10,11,6,7,8,9,12,13)])
# idx <- c(14,2,3,4,5,10,11,6,7,8,9,12,13)
# 
# pdf(file.path(cwd, "figures/NGS3660_somatic_mutations_AFs.pdf"), width=10, height=15)
# pheatmap(d4plot[,idx], scale = "none", cluster_cols = FALSE, fontsize_col = 8, angle_col=45)
# dev.off()
```

## Generate two plots 1) AF accross the interested mutations 

```{r}

r_effect = r%>%
  dplyr::mutate(gene = gsub(",.*","", effects.merged))%>%
  dplyr::mutate(gene = gsub("NA \\| ","", gene))%>%
  dplyr::mutate(isMissense = case_when(grepl("missense", effects.merged) ~ "Missense", 
                                       T ~ ""))%>%
  dplyr::mutate(isNMD = case_when(grepl("NMD_", effects.merged) ~ "NMD", 
                                       T ~ ""))%>%
  dplyr::mutate(isframeshift = case_when(grepl("frameshift_variant", effects.merged) ~ "Frameshift",   #### no one is frameshift
                                       T ~ ""))%>%
  dplyr::mutate(isStartlost = case_when(grepl("start_lost", effects.merged) ~ "Startlost",    ### no one is start loss
                                       T ~ ""))%>%
  dplyr::mutate(isStopgain = case_when(grepl("stop_gained", effects.merged) ~ "Stopgained", 
                                       T ~ ""))%>%
  dplyr::select(-isStartlost)%>%
  dplyr::select(-isframeshift)%>%
  dplyr::mutate(alltype = paste(isMissense, isNMD, isStopgain, sep = ""))%>%
  dplyr::select(ID, gene, alltype)%>%
  unique()


r_effect_sort = data.frame(ID = rownames(freqs.sel.bind), stringsAsFactors = F)%>%
  dplyr::left_join(r_effect, by = "ID")

freq_tile = freqs.sel.bind

rownames(freq_tile) =r_effect_sort$gene

ta = HeatmapAnnotation(passage =c("P1","P2", "P3","P4","P5"),
                       col = list(passage = c("P1" = "#2D8686", "P2"= "#592D86",
                                              "P3" = "darkorange", "P4"= "darkgreen",
                                              "P5" = "#862D2D")))


la = rowAnnotation(
  mutation_type = r_effect_sort$alltype,
  col = list(
    mutation_type = c(
      "Missense" = "seagreen",
      "MissenseNMD" = "gold",
      "NMD" = "darkorange",
      "Stopgained" = "maroon"
    )
  )
)




naToZero_dist = function(x,y) {
  
  x[is.na(x)] = 0 
  y[is.na(y)] = 0 

  return(1-cor(x,y))
}




hei = unit(square_cal(nrow(freq_tile))*40, "cm")
col_fun_expr = colorRamp2(c(0, 1), c("white", "red"))


h1 = Heatmap(freq_tile,
             name = "Allele Frequency",
        col = col_fun_expr,
        row_names_side = "left",
        height =  hei,
        width = hei*(ncol(freq_tile)/nrow(freq_tile)),
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 6),
        rect_gp = gpar(col = "black", lwd = 0.5),
        left_annotation = la, 
        top_annotation = ta, 
        cluster_rows = T, 
        clustering_distance_rows = naToZero_dist,
        cluster_columns = F)



read_tile = freqs.sel.reads.bind

rownames(read_tile) =r_effect_sort$gene

ta = HeatmapAnnotation(passage =c("P1","P2", "P3","P4","P5"),
                       col = list(passage = c("P1" = "#2D8686", "P2"= "#592D86",
                                              "P3" = "darkorange", "P4"= "darkgreen",
                                              "P5" = "#862D2D")))


la = rowAnnotation(
  mutation_type = r_effect_sort$alltype,
  col = list(
    mutation_type = c(
      "Missense" = "seagreen",
      "MissenseNMD" = "gold",
      "NMD" = "darkorange",
      "Stopgained" = "maroon"
    )
  )
)






hei = unit(square_cal(nrow(read_tile))*40, "cm")
col_fun_expr = colorRamp2(c(0, 200), c("white", "black"))


h2 = Heatmap(read_tile,
             name = "total reads",
        col = col_fun_expr,
        row_names_side = "left",
        height =  hei,
        width = hei*(ncol(read_tile)/nrow(read_tile)),
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 6),
        rect_gp = gpar(col = "black", lwd = 0.5),
        #left_annotation = la, 
        #top_annotation = ta, 
        cluster_rows = T, 
        clustering_distance_rows = naToZero_dist,
        cluster_columns = F)


pdf("/gstore/project/crc_metastasis/revision/P1P2P3P4P5_WES/figures/mut_freq_heatmap_noAF_P12345_20250529.pdf", height = 18, width = 5, useDingbats = F)
 h1+ h2
 dev.off()
 
 
 
 saveRDS(r_effect_sort, "/gstore/project/crc_metastasis/revision/P1P2P3P4P5_WES/r_effect_sort_20250529.rds")
 saveRDS(freq_tile, "/gstore/project/crc_metastasis/revision/P1P2P3P4P5_WES/freq_tile_20250529.rds")
 saveRDS(read_tile, "/gstore/project/crc_metastasis/revision/P1P2P3P4P5_WES/read_tile_20250529.rds")

```


## from the above heatmap remove rows where the max frequency is less than 0.4

```{r}



max_freq = apply(freq_tile, 1, max)
which_high = which(max_freq >0.39)
high_freq_tile = freq_tile[which_high, ]
high_read_tile = read_tile[which_high,]


hei = unit(square_cal(nrow(high_freq_tile))*40, "cm")
col_fun_expr = colorRamp2(c(0, 1), c("white", "red"))

la = rowAnnotation(
  mutation_type = r_effect_sort$alltype[which_high],
  col = list(
    mutation_type = c(
      "Missense" = "seagreen",
      "MissenseNMD" = "gold",
      "NMD" = "darkorange",
      "Stopgained" = "maroon"
    )
  )
)


h1 = Heatmap(high_freq_tile,
             name = "Allele Frequency",
        col = col_fun_expr,
        row_names_side = "left",
        height =  hei,
        width = hei*(ncol(high_freq_tile)/nrow(high_freq_tile)),
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 6),
        rect_gp = gpar(col = "black", lwd = 0.5),
        left_annotation = la, 
        top_annotation = ta, 
        cluster_rows = T, 
        clustering_distance_rows = naToZero_dist,
        cluster_columns = F)





hei = unit(square_cal(nrow(high_read_tile))*40, "cm")
col_fun_expr = colorRamp2(c(0, 200), c("white", "black"))


h2 = Heatmap(high_read_tile,
             name = "total reads",
        col = col_fun_expr,
        row_names_side = "left",
        height =  hei,
        width = hei*(ncol(high_read_tile)/nrow(high_read_tile)),
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 6),
        rect_gp = gpar(col = "black", lwd = 0.5),
        #left_annotation = la, 
        #top_annotation = ta, 
        cluster_rows = T, 
        clustering_distance_rows = naToZero_dist,
        cluster_columns = F)


pdf("/gstore/project/crc_metastasis/revision/P1P2P3P4P5_WES/figures/mut_high_freq_heatmap_noAF_P12345_20250529.pdf", height = 12, width = 5, useDingbats = F)
 h1+ h2
 dev.off()

```




20250721 
create another version without AF, just the categories of mutations 


```{r}


max_freq = apply(freq_tile, 1, max)
which_high = which(max_freq >0.39)
high_freq_tile = freq_tile[which_high, ]
high_read_tile = read_tile[which_high,]

high_r_effect_sort = r_effect_sort[which_high,]

high_tile = cbind(high_r_effect_sort, high_freq_tile)%>%
  dplyr::select(gene, SAM24400842,SAM24400845, SAM24488704, SAM24488703, SAM24400853, alltype)%>%
    dplyr::mutate(SAM24400842 = case_when(SAM24400842==0 ~ NA, 
                          T ~ alltype))%>%
  dplyr::mutate(SAM24400845 = case_when(SAM24400845==0 ~ NA, 
                          T ~ alltype))%>%
  dplyr::mutate(SAM24488704 = case_when(SAM24488704==0 ~ NA, 
                          T ~ alltype))%>%
  dplyr::mutate(SAM24488703 = case_when(SAM24488703==0 ~ NA, 
                          T ~ alltype))%>%
  dplyr::mutate(SAM24400853 = case_when(SAM24400853==0 ~ NA, 
                          T ~ alltype))%>%
  dplyr::select(-alltype)




tm=as.matrix(high_tile[, 2:6])
rownames(tm) = high_tile$gene

ta = HeatmapAnnotation(passage =c("P1","P2", "P3","P4","P5"),
                       col = list(passage = c("P1" = "#2D8686", "P2"= "#592D86",
                                              "P3" = "darkorange", "P4"= "darkgreen",
                                              "P5" = "#862D2D")))


dist_mut = function(x, y) {
  
  x[is.na(x)] = 0
  x[which(x == "Missense")] = 1
  x[which(x == "MissenseNMD")] = 2
  x[which(x == "NMD")] = 3
  x[which(x == "Stopgained")] = 4
  x= as.numeric(x)

  y[is.na(y)] = 0
  y[which(y == "Missense")] = 1
  y[which(y == "MissenseNMD")] = 2
  y[which(y == "NMD")] = 3
  y[which(y == "Stopgained")] = 4
  y= as.numeric(y)

  
    sqrt(sum((x - y)^2))
}


hei = unit(square_cal(nrow(tm))*40, "cm")

hm = Heatmap(tm, 
        name = "mutations", 
        col = structure(c("seagreen","gold","darkorange","maroon"),
                                                names = c("Missense","MissenseNMD","NMD","Stopgained")),
        height = hei, 
        width = hei*(ncol(tm)/nrow(tm)), 
        rect_gp = gpar(col = "black", lwd = 0.5),
        row_names_side = "left",
        row_names_gp = gpar(fontsize = 10),
        column_names_gp = gpar(fontsize = 6),
        top_annotation = ta, 
        clustering_distance_rows = dist_mut, 
        cluster_columns = F
    )


pdf("/gstore/project/crc_metastasis/revision/P1P2P3P4P5_WES/figures/mut_high_freq_heatmap_categorical_P12345_20250721.pdf", height = 12, width = 5, useDingbats = F)
 hm
 dev.off()


```














functions 


```{r}

process.lofreq <- function(x){

  d.r <- rowRanges(x) %>% as.data.frame()
  d.i <- info(x) %>% as.data.frame()
  d.r$ALT <- unlist(sapply(d.r$ALT, function(x){paste(as.character(x),collapse=",")}))
  d.r <- cbind(d.r, d.i[,c("DP","AF")])

  d.csq.l <- lapply(seq_along(d.i$CSQ), function(i){
    #message(i)
    csq <- d.i$CSQ[[i]]
    d.csq <- do.call(rbind,lapply(strsplit(csq, "\\|"), function(x){x[1:40]}))
    d.csq.str <- paste(apply(unique(d.csq[,c(4,2,17,16,14,15,7,26),drop=FALSE]),1, paste, collapse=","), collapse=" | ")
  })

  d.r$CSQ <- unlist(d.csq.l)
  d.r <- data.frame(ID=rownames(d.r), d.r)
  return(d.r)
}

process.strelka <- function(x){
  d.r <- rowRanges(x) %>% as.data.frame() %>% mutate(ID=rownames(.))

  d.i <- info(x) %>% as.data.frame()
  d.g <- geno(x)
  d.r$ALT <- unlist(sapply(d.r$ALT, function(x){paste(as.character(x),collapse=",")}))


  AF.l <- lapply(1:nrow(d.r), function(i){

    nuc.ref <- d.r[i, "REF"]
    nuc.alt <- d.r[i, "ALT"]

    cnt.ref <- d.g[[sprintf("%sU",nuc.ref)]][i, 2, 1]
    d <- lapply(unlist(strsplit(nuc.alt, ",")), function(x){
      cnt.alt <- d.g[[sprintf("%sU",x)]][i, 2, 1]
      AF <- cnt.alt/(cnt.alt + cnt.ref)
      DP <- (cnt.alt + cnt.ref)
      return(data.frame(ID=d.r$ID[i],REF=nuc.ref, ALT=x, DP=DP, AF=AF))
    })
    d.bind <- do.call(rbind, d)
    # cnt.alt <- d.g[[sprintf("%sU",nuc.alt)]][i, 2, 1]
    # 
    # AF <- cnt.alt/(cnt.alt + cnt.ref)
    # DP <- (cnt.alt + cnt.ref)
    # return(c(DP=DP,AF=AF))
  })
  d.r.merge <- merge(d.r, do.call(rbind, AF.l), by="ID") %>% mutate(REF=REF.y, ALT=ALT.y) %>% dplyr::select(-c("REF.x", "ALT.x", "REF.y", "ALT.y"))
  d.r.merge <- d.r.merge[,c(1:7,12,13,8:11)]

  d.csq.l <- lapply(seq_along(d.i$CSQ), function(i){
    #message(i)
    csq <- d.i$CSQ[[i]]
    d.csq <- do.call(rbind,lapply(strsplit(csq, "\\|"), function(x){x[1:40]}))
    d.csq.str <- paste(apply(unique(d.csq[,c(4,2,17,16,14,15,7,26),drop=FALSE]),1, paste, collapse=","), collapse=" | ")
  })
  names(d.csq.l) <- rownames(d.i)

  d.r.merge$CSQ <- unlist(d.csq.l[d.r.merge$ID])
  d.r.merge$ID <-  sprintf("%s:%s_%s/%s", d.r.merge$seqnames, d.r.merge$start, d.r.merge$REF, d.r.merge$ALT)


  return(d.r.merge)
  #d.r <- data.frame(ID=rownames(d.r), d.r)

}


process.mutect2 <- function(x, pass.only=TRUE){
  library(parallel)

  if(pass.only){
    x <- x[rowRanges(x)$FILTER == "PASS"]
  }

  d.r <- rowRanges(x) %>% as.data.frame()
  d.i <- info(x) %>% as.data.frame()
  d.g <- geno(x) #cbind(d.r, d.i[,c("DP","AF")])
  d.r$ALT <- unlist(sapply(d.r$ALT, function(x){paste(as.character(x),collapse=",")}))
  d.r$DP <- unlist(d.g$DP[,2])
  #d.r$AF <- unlist(sapply(d.g$AF[,2], function(x){paste(as.character(x),collapse=",")}))
  d.r$AF <- unlist(sapply(d.g$AF[,2], function(x){max(x)})) # only select max AF

  d.csq.l <- mclapply(seq_along(d.i$CSQ), function(i){
    message(i)
    csq <- d.i$CSQ[[i]]
    d.csq <- do.call(rbind,lapply(strsplit(csq, "\\|"), function(x){x[1:40]}))
    d.csq.str <- paste(apply(unique(d.csq[,c(4,2,17,16,14,15,7,26),drop=FALSE]),1, paste, collapse=","), collapse=" | ")
  }, mc.cores=6)

  d.r$CSQ <- unlist(d.csq.l)
  d.r <- data.frame(ID=rownames(d.r), d.r)
  return(d.r)
}



###

getAllelesFromBamSimple <- function(positions, bams){

  nonZeroOnly = FALSE; mutatedSitesOnly = FALSE
  pParam <- Rsamtools::PileupParam(distinguish_nucleotides = TRUE,
                                   distinguish_strands = FALSE, max_depth = 5000, min_base_quality = 25,
                                   min_nucleotide_depth = 0, include_deletions = FALSE,
  )
  sbp <- Rsamtools::ScanBamParam(which = positions)


  res <- lapply(bams, function(x) Rsamtools::pileup(x, scanBamParam = sbp,pileupParam = pParam))

  purrr::walk(seq_along(res), function(x) if (nrow(res[[x]]) == 0)
    warning(paste0("No mapped reads for given positions for ",
                   names(res[x]), " sample.")))
  if (nonZeroOnly == TRUE & length(res) == 0)
    stop("No mapped reads for any samples.")
  res <- lapply(res, function(x) x <- x %>% dplyr::mutate_if(is.factor,
                                                             as.character))
  if (nonZeroOnly == FALSE) {
    res2 <- data.frame(seqnames = as.character(GenomicRanges::seqnames(positions)),
                       pos = GenomicRanges::ranges(positions)@start, nucleotide = NA,
                       count = 0, stringsAsFactors = FALSE)
    res <- suppressWarnings(lapply(res, function(x) x <- dplyr::left_join(res2,
                                                                          x, by = c("seqnames", "pos")) %>% dplyr::mutate(nucleotide = ifelse(is.na(nucleotide.y),
                                                                                                                                              NA, as.character(nucleotide.y)), count = ifelse(is.na(count.y),
                                                                                                                                                                                              NA, count.y)) %>% dplyr::select(seqnames, pos, nucleotide,

                       count)))
  }
  res <- res[lapply(res, function(x) nrow(x)) > 0]
  res <- lapply(res, function(x) split(x, paste0(x$seqnames,
                                                 ":", x$pos)))
  alleles <- lapply(res, function(x) lapply(x, function(y) y$nucleotide))
  readsAlleles <- lapply(res, function(x) lapply(x, function(y) ifelse(is.na(as.numeric(y$count)),
                                                                       1, as.numeric(y$count))))
  SNV <- Map(function(i, j) Map(function(x, y) rep(x, y),
                                i, j), alleles, readsAlleles)
  SNV[which(sapply(SNV, length) == 0)] <- list(rep(list(NULL),
                                                   1))

  df <- lapply(SNV, function(x) lapply(seq(length(x)), function(y) {
    data.frame(position = names(x[y]), numberOfReads = length(stats::na.omit(x[[y]])),
               fracA = round(sum(x[[y]] == "A")/length(x[[y]]), 3),
               fracC = round(sum(x[[y]] == "C")/length(x[[y]]), 3),
               fracG = round(sum(x[[y]] == "G")/length(x[[y]]), 3),
               fracT = round(sum(x[[y]] == "T")/length(x[[y]]), 3))
  }))

  if (mutatedSitesOnly == TRUE) {
    df <- lapply(df, function(x) lapply(x, function(y) {
      if (!is.na(y[[paste0("frac", y[["refAllele"]])]]) &
          y[[paste0("frac", y[["refAllele"]])]] == 1) {
        y <- NULL
        y
      }
      else y
    }))
    df <- lapply(df, function(x) x[lapply(x, length) > 0])
  }
  df <- suppressWarnings(lapply(df, function(x) dplyr::bind_rows(x)))

  return(df)
}



dist_mut = function(x, y) {
  
  x[is.na(x)] = 0
  x[which(x == "Missense")] = 1
  x[which(x == "MissenseNMD")] = 2
  x[which(x == "NMD")] = 3
  x[which(x == "Stopgained")] = 4
  x= as.numeric(x)

  y[is.na(y)] = 0
  y[which(y == "Missense")] = 1
  y[which(y == "MissenseNMD")] = 2
  y[which(y == "NMD")] = 3
  y[which(y == "Stopgained")] = 4
  y= as.numeric(y)

  
    sqrt(sum((x - y)^2))
}

square_cal = function(n)
{
  if(n > 10)
  {
      c = ceiling(log10(n))
  p = n/(10**c)
  }else{
      c = ceiling(log10(n))
  p = n/(100**c)

  }

  if(n == 1)
  {
    p = 0.01
  }

  return(p)

}


```

test


```{r}
getAllelesFromBamSimple(posExample[1:2], files.bam["SAM24400842"])
```






